---
title: "University Ranking Analysis"
author: |
  Group 5  
  22-46624-1  
  22-46667-1  
  22-46342-1  
  22-47180-1
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    df_print: paged
    toc: true
    toc_depth: 2
    number_sections: false
  pdf_document:
    toc: true
    toc_depth: 2
---



## Data Acquisition and Initial Exploration: 
Import necessary library and load the dataset
```{r}


library(ggplot2)
library(GGally)
library(dplyr)
library(ggcorrplot)
library(RColorBrewer)
library(tidyverse)
library(psych)
library(patchwork)


url <- "https://raw.githubusercontent.com/iamrakiib/Predicting-University-Score-Based-on-Key-Indicators/main/cwurData.csv"
df <- read.csv(url)
```

An initial inspection to understand its structure and variable types.
```{r}
#view structure of the dataset
print(str(df))

```



```{r}
# View first few rows
print(head(df))

```

```{r}
# View column names
print(colnames(df))
```
```{r}
# Summary statistics
print(summary(df))
```

## Data Cleaning and Wrangling
Identify and handle missing values, duplicates, and any other data quality issues.

```{r}
#check Missing Values Per Column
print(colSums(is.na(df)))

```
```{r}
#check missing values in the entire dataset
print(sum(is.na(df)))
```
```{r}
#check which rows have missing values
print(sum(!complete.cases(df)))

```
```{r}
#check if there is any duplicate rows
print(sum(duplicated(df)))
```


```{r}
library(e1071)
skew_broad_impact <- skewness(df$broad_impact, na.rm = TRUE) # na.rm=true, missing value bad diye count kore
cat("Skewness of 'broad_impact':", skew_broad_impact, "\n") 
```
```{r}
# Fill missing values in 'broad_impact' with the mean as skew is symmetric
df$broad_impact[is.na(df$broad_impact)] <- mean(df$broad_impact, na.rm = TRUE)
```


```{r}
#after filling with mean check is there any missing value remains in broad impact
print(sum(is.na(df$broad_impact)))
```

```{r}
#after filling with mean check is there any missing value remains in any column
print(colSums(is.na(df)))
```
```{r}
#check if the missing value is properly filled
print(str(df))
```


```{r}
# Combined university count extraction and printing
university_count_in_2015 <- nrow(df[df$year == 2015, ])
university_count_before_2015 <- nrow(df[df$year < 2015, ])
university_count_in_2014 <- nrow(df[df$year == 2014, ])
university_count_in_2013 <- nrow(df[df$year == 2013, ])
university_count_in_2012 <- nrow(df[df$year == 2012, ])

# Print all counts
print(paste("Universities in 2015:", university_count_in_2015))
print(paste("Universities before 2015:", university_count_before_2015))
print(paste("Universities in 2014:", university_count_in_2014))
print(paste("Universities in 2013:", university_count_in_2013))
print(paste("Universities in 2012:", university_count_in_2012))

```



```{r}
# Nominal (unordered categories)
df$country <- as.factor(df$country)
df$institution <- as.factor(df$institution)

# Ordinal (ordered categories - optional)
# For example, creating a rank group (optional)


df <- df %>%
  group_by(year) %>%
  arrange(year, world_rank) %>%
  mutate(year_rank = row_number()) %>%
  ungroup() %>%
  mutate(rank_group = cut(year_rank,
                          breaks = c(0, 100, 500, Inf),
                          labels = c("Top Tier (1-100)", "Mid Tier (101-500)", "Lower Tier (501+)"),
                          right = TRUE)) %>%
  mutate(rank_group = as.factor(rank_group))

print(head(df))
```

```{r}


summary_stats <- df %>%
  group_by(year, rank_group) %>%
  summarise(
    avg_score = mean(score, na.rm = TRUE),
    avg_publications = mean(publications, na.rm = TRUE),
    influence=mean(influence, na.rm = TRUE),
    count = n()
  ) %>%
  arrange(year, rank_group)

print(summary_stats)

```


## Exploratory Data Analysis(EDA)
Key Descriptive Statistics
```{r}
# Summary statistics for numerical variables
summary(df)

# More detailed stats (mean, median, sd, min, max, quartiles)
library(psych)
describe(df[sapply(df, is.numeric)])

```

Distributions of Key Variables
Histogram of Numerical Features

```{r}

# Identify numeric columns
numeric_cols <- names(df)[sapply(df, is.numeric)]

# Loop through numeric columns and plot histograms
for (col in numeric_cols) {
  p <- ggplot(df, aes_string(x = col)) +
    geom_histogram(fill = "skyblue", color = "black", bins = 30) +
    labs(title = paste("Distribution of", col), x = col, y = "University Count") +
    theme_minimal()
  
  print(p)
}

```



```{r}
ggplot(df, aes(x = score)) +
  geom_histogram(binwidth = 2, fill = "#2c7bb6", color = "black", alpha = 0.8) +
  labs(title = "Distribution of University Scores",
       x = "Score",
       y = "University Count in this Interval") +
  theme_minimal()

# 2. Distribution of Publications
ggplot(df, aes(x = publications)) +
  geom_histogram(binwidth = 50, fill = "#d95f02", color = "black", alpha = 0.8) +
  labs(title = "Distribution of Publications",
       x = "Publications",
       y = "University Count in this Interval") +
  theme_minimal()

# 3. Distribution of Citations
ggplot(df, aes(x = citations)) +
  geom_histogram(binwidth = 50, fill = "#7570b3", color = "black", alpha = 0.8) +
  labs(title = "Distribution of Citations",
       x = "Citations",
       y = "University Count in this Interval") +
  theme_minimal()

# 4. Distribution of Influence
ggplot(df, aes(x = influence)) +
  geom_histogram(binwidth = 2, fill = "#1b9e77", color = "black", alpha = 0.8) +
  labs(title = "Distribution of Influence",
       x = "Influence",
       y = "University Count in this Interval") +
  theme_minimal()

# 5. Distribution of World Rank
ggplot(df, aes(x = broad_impact)) +
  geom_histogram(binwidth = 50, fill = "#e7298a", color = "black", alpha = 0.8) +
  labs(title = "Distribution of World Rank",
       x = "Broad Impact",
       y = "University Count in this Interval") +
  theme_minimal()

```
```{r}
library(dplyr)

# Suppose your data frame is df and the column is broad_impact
specific_value <- 496.6995
  # replace with the value you want

count_specific <- df %>%
  filter(broad_impact == specific_value) %>%
  summarise(count = n())

print(count_specific)


```
```{r}
library(dplyr)

# Suppose your data frame is called df and the column is broad_impact
# Set your desired range
lower <- 480
upper <- 522

# Count universities in that range
count_in_range <- df %>%
  filter(broad_impact >= lower & broad_impact <= upper) %>%
  summarise(count = n())

print(count_in_range)

```



Scatterplot for numeric column for all three tiers 

```{r}

# Create tier column if not already done
df <- df %>%
  mutate(tier = case_when(
    world_rank <= 100 ~ "Top Tier",
    world_rank <= 500 ~ "Mid Tier",
    TRUE ~ "Low Tier"
  ))

# Define colors for tiers
tier_colors <- c("Top Tier" = "blue", "Mid Tier" = "orange", "Low Tier" = "green")

# Example for Score vs Influence (repeat similarly for other plots)
ggplot(df, aes(x = score, y = influence, color = tier, group = tier)) +
  geom_point(alpha = 0.6) +                # points only, no lines between points
  scale_color_manual(values = tier_colors) +
  labs(title = "Score vs Influence by Tier",
       x = "Score", y = "Influence", color = "Tier") +
  theme_minimal()


# 2. Score vs Citations
ggplot(df, aes(x = score, y = citations, color = tier, group = tier)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = tier_colors) +
  labs(title = "Score vs Citations by Tier",
       x = "Score", y = "Citations", color = "Tier") +
  theme_minimal()

# 3. Score vs Publications
ggplot(df, aes(x = score, y = publications, color = tier, group = tier)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = tier_colors) +
  labs(title = "Score vs Publications by Tier",
       x = "Score", y = "Publications", color = "Tier") +
  theme_minimal()

# 4. Score vs Broad Impact
ggplot(df, aes(x = score, y = broad_impact, color = tier, group = tier)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = tier_colors) +
  labs(title = "Score vs Broad Impact by Tier",
       x = "Score", y = "Broad Impact", color = "Tier") +
  theme_minimal()

# 5. Score vs World Rank (reverse y axis)
ggplot(df, aes(x = score, y = world_rank, color = tier, group = tier)) +
  geom_point(alpha = 0.6) +
  scale_y_reverse() +
  scale_color_manual(values = tier_colors) +
  labs(title = "Score vs World Rank by Tier",
       x = "Score", y = "World Rank (Lower is Better)", color = "Tier") +
  theme_minimal()

# 6. Citations vs World Rank (reverse y axis)
ggplot(df, aes(x = citations, y = world_rank, color = tier, group = tier)) +
  geom_point(alpha = 0.6) +
  scale_y_reverse() +
  scale_color_manual(values = tier_colors) +
  labs(title = "Citations vs World Rank by Tier",
       x = "Citations", y = "World Rank (Lower is Better)", color = "Tier") +
  theme_minimal()
```
```{r}

# Filter for universities with world_rank between 0 and 250
filtered_df <- df %>%
  filter(world_rank >= 750, world_rank <= 1000)

# Total citation count in that rank range
total_citations <- sum(filtered_df$citations, na.rm = TRUE)

# Count of universities with citation count between 0 and 200 in that rank range
count_citations_0_200 <- filtered_df %>%
  filter(citations >=200, citations <= 400) %>%
  summarise(count = n()) %>%
  pull(count)

list(
  total_citations = total_citations,
  count_citations_0_200 = count_citations_0_200
)

```

Correlation Heatmap

```{r}
# Step 1: Select numeric columns
numeric_df <- df[sapply(df, is.numeric)]

# Step 2: Compute correlation matrix (excluding NAs)
cor_matrix <- cor(df[, sapply(df, is.numeric)], use = "complete.obs")


ggcorrplot(cor_matrix,
           method = "square",
           type = "lower",
           lab = TRUE,
           lab_size = 2.5,
           colors = c("#B2182B", "white", "#2166BC"),
           title = "University Metrics Correlation",
           ggtheme = ggplot2::theme_minimal()) +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))

```


Boxplots to Detect Outliers by Rank Group
This shows how metrics differ across rank tiers.

```{r}
ggplot(df, aes(x = rank_group, y = score, fill = rank_group)) +
  geom_boxplot() +
  labs(title = "Score Distribution by Rank Tier", x = "Rank Tier", y = "Score") +
  theme_minimal()

```



```{r}

# Parameters: specify the year and country you want to filter
target_year <- 2015
target_country <- "USA"

# Filter dataset by year and country, then count unique universities
university_count <- df %>%
  filter(year == target_year, country == target_country) %>%
  summarise(count = n_distinct(institution)) %>%
  pull(count)

print(university_count)

```


Line Plot
```{r}


df <- df %>%
  mutate(tier = case_when(
    world_rank <= 100 ~ "Top Tier",
    world_rank <= 500 ~ "Mid Tier",
    TRUE ~ "Low Tier"
  ))

tier_colors <- c("Top Tier" = "blue", "Mid Tier" = "orange", "Low Tier" = "green")

# 1. Average Score by Year and Tier
ggplot(df %>% group_by(year, tier) %>% summarise(avg_score = mean(score, na.rm = TRUE), .groups = "drop"),
       aes(x = year, y = avg_score, color = tier)) +
  geom_line(size = 1.2) + geom_point(size = 3) +
  scale_color_manual(values = tier_colors) +
  scale_x_continuous(breaks = unique(df$year)) +
  labs(title = "Average Score Over Years", x = "Year", y = "Average Score") +
  theme_minimal()

ggplot(df %>% 
         group_by(year, tier) %>% 
         summarise(avg_citations = mean(citations, na.rm = TRUE), .groups = "drop"),
       aes(x = year, y = avg_citations, color = tier)) +
  geom_line(size = 1.2) + 
  geom_point(size = 3) +
  scale_color_manual(values = tier_colors) +
  scale_x_continuous(breaks = unique(df$year)) +
  scale_y_continuous(breaks = seq(0, max(df$citations, na.rm = TRUE) + 100, by = 100)) +
  labs(title = "Average Citations Over Years", x = "Year", y = "Average Citations") +
  theme_minimal()


# 3. Average Publications by Year and Tier
ggplot(df %>% group_by(year, tier) %>% summarise(avg_publications = mean(publications, na.rm = TRUE), .groups = "drop"),
       aes(x = year, y = avg_publications, color = tier)) +
  geom_line(size = 1.2) + geom_point(size = 3) +
  scale_color_manual(values = tier_colors) +
  scale_x_continuous(breaks = unique(df$year)) +
  scale_y_continuous(breaks = seq(0, max(df$publications, na.rm = TRUE) + 100, by = 100)) +

  labs(title = "Average Publications Over Years", x = "Year", y = "Average Publications") +
  theme_minimal()

# 4. Average Broad Impact by Year and Tier
ggplot(df %>% group_by(year, tier) %>% summarise(avg_broad_impact = mean(broad_impact, na.rm = TRUE), .groups = "drop"),
       aes(x = year, y = avg_broad_impact, color = tier)) +
  geom_line(size = 1.2) + geom_point(size = 3) +
  scale_color_manual(values = tier_colors) +
  scale_x_continuous(breaks = unique(df$year)) +
  labs(title = "Average Broad Impact Over Years", x = "Year", y = "Average Broad Impact") +
  theme_minimal()

ggplot(df %>%
         group_by(year, tier) %>%
         summarise(avg_quality_edu = mean(quality_of_education, na.rm = TRUE), .groups = "drop"),
       aes(x = year, y = avg_quality_edu, color = tier)) +
  geom_line(size = 1.2) + geom_point(size = 3) +
  scale_color_manual(values = tier_colors) +
  scale_x_continuous(breaks = unique(df$year)) +
  labs(title = "Average Quality of Education Over Years by Tier",
       x = "Year", y = "Quality of Education Score") +
  theme_minimal()

ggplot(df %>%
         group_by(year, tier) %>%
         summarise(avg_influence = mean(influence, na.rm = TRUE), .groups = "drop"),
       aes(x = year, y = avg_influence, color = tier)) +
  geom_line(size = 1.2) + geom_point(size = 3) +
  scale_color_manual(values = tier_colors) +
  scale_x_continuous(breaks = unique(df$year)) +
  labs(title = "Average Influence Over Years by Tier",
       x = "Year", y = "Influence Score") +
  theme_minimal()



```

```{r}
# Select top 6 countries across all years (overall)
top_countries <- df %>%
  group_by(country) %>%
  summarise(total = n()) %>%
  top_n(5, total) %>%
  pull(country)

# Filter only those countries
df_filtered <- df %>%
  filter(country %in% top_countries)

# Count universities per year per country
country_year_trend <- df_filtered %>%
  group_by(year, country) %>%
  summarise(university_count = n(), .groups = "drop")

# Plot line chart
ggplot(country_year_trend, aes(x = year, y = university_count, color = country)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(title = "Trends of Ranked Universities per Year (Top Countries)",
       x = "Year",
       y = "Number of Universities") +
  theme_minimal()

```
```{r}
df %>%
  filter(country == "China", 
         year %in% 2012:2015) %>%
  count(year, name = "university_count")
```



Boxplot of different attributes

```{r}
df %>%
  mutate(pub_tertile = cut_number(publications, 3, labels = c("Low", "Medium", "High"))) %>%
  ggplot(aes(x = pub_tertile, y = influence, fill = pub_tertile)) +
  geom_boxplot() +
  labs(title = "Research Influence by Publication Volume", 
       x = "Publication Tertile", y = "Influence Score") +
  scale_fill_brewer(palette = "Pastel1") +
  theme_minimal()

df %>%
  mutate(year = factor(year)) %>%
  ggplot(aes(x = year, y = broad_impact, fill = year)) +
  geom_boxplot() +
  labs(title = "Broad Impact Trends (2012-2015)", 
       x = "Year", y = "Broad Impact Score") +
  scale_fill_viridis_d() +
  theme_minimal()

ggplot(df, aes(x = factor(year), y = score)) +
  geom_boxplot(fill = "lightblue", outlier.color = "red") +
  labs(title = "University Scores Distribution (2012-2015)", 
       x = "Year", y = "Score") +
  theme_minimal()


ggplot(df, aes(x = factor(year), y = citations)) +
  geom_boxplot(fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Citations Distribution by Year",
       x = "Year",
       y = "Citations") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0),
    axis.title = element_text(face = "bold")
  )


ggplot(df, aes(x = rank_group, y = patents, fill = rank_group)) +
  geom_boxplot() +
  labs(title = "Patents Filed by University Tier", 
       x = "Rank Group", y = "Patents") +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal()

# Get top 5 countries by number of universities
top5_countries <- df %>%
  group_by(country) %>%
  summarise(count = n(), .groups = "drop") %>%
  arrange(desc(count)) %>%
  slice_head(n = 5) %>%
  pull(country)

# Filter and plot
df %>%
  filter(country %in% top5_countries) %>%
  ggplot(aes(x = country, y = score, fill = country)) +
  geom_boxplot() +
  labs(title = "University Scores by Top 5 Countries",
       x = "Country",
       y = "Score") +
  scale_fill_brewer(palette = "Set3") +
  theme_minimal()


```
Pairplot of different numeric attributes

```{r}
# Ensure rank_group is a factor and assign numeric colors
df$rank_group <- as.factor(df$rank_group)
colors <- as.numeric(df$rank_group)  # This will convert factor levels to 1, 2, 3

# Select relevant columns
selected_data <- df[, c("score", "publications", "citations")]

# Create pair plot
pairs(selected_data,
      col = colors,
      pch = 19,
      labels = c("Score", "Publications", "Citations"),  # custom column names
      main = "Pair Plot of University Features Colored by Rank Group")

```




```{r}
# Ensure rank_group is a factor and assign numeric colors
df$rank_group <- as.factor(df$rank_group)
colors <- as.numeric(df$rank_group)  # Factor levels → numeric colors

# Select relevant numeric columns
selected_data <- df[, c("score", "publications", "citations", 
                        "influence", "broad_impact", "patents")]

# Create pair plot without tier labels
pairs(selected_data,
      col = colors,
      pch = 19,
      labels = c("Score", "Publications", "Citations","influence", "broad_impact", "patents"),
      main = "Pair Plot of University Features")

```


Barplot
```{r}

ggplot(summary_stats, aes(x = year, y = count, fill = rank_group)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Number of Universities by Rank Group and Year",
       x = "Year", y = "Count") +
  theme_minimal()

```
```{r}

library(dplyr)
library(ggplot2)

# Assuming your data is in `df` with columns: `country`, `year`, `university_count` (or similar)

# Get top 5 countries per year
top_countries_each_year <- df %>%
  group_by(year, country) %>%
  summarise(university_count = n(), .groups = "drop") %>%
  group_by(year) %>%
  slice_max(order_by = university_count, n = 5) %>%
  ungroup()

# Plot
ggplot(top_countries_each_year, 
       aes(x = reorder(country, university_count), 
           y = university_count, 
           fill = country)) +  # Color by country instead of year
  geom_col(position = "dodge") +
  facet_wrap(~ year, scales = "free") +  # Separate plots per year
  coord_flip() +  # Horizontal bars
  labs(title = "Top 5 Countries by Number of Ranked Universities (Each Year)",
       x = "Country",
       y = "Number of Universities",
       fill = "Country") +
  theme_minimal() +
  theme(legend.position = "none")  # Remove legend (redundant since country is on y-axis)
```

```{r}


# Filter data for 2012-2015 and compute top 5 countries by avg score per year
df_yearly_avg <- df %>%
  filter(year %in% 2012:2015) %>%  # Keep only 2012-2015
  group_by(year, country) %>%
  summarise(avg_score = mean(score, na.rm = TRUE), .groups = "drop") %>%
  group_by(year) %>%
  arrange(year, desc(avg_score)) %>%  # Sort by year and score (high to low)
  slice(1:5) %>%  # Take top 5 per year
  ungroup()

# Plot faceted by year
ggplot(df_yearly_avg, 
       aes(x = reorder(country, avg_score), 
           y = avg_score, 
           fill = country)) +
  geom_bar(stat = "identity") +
  coord_flip() +  # Horizontal bars
  facet_wrap(~ year, ncol = 2, scales = "free_y") +  # 2x2 grid of plots
  labs(title = "Top 5 Countries by Average Score (2012-2015)",
       x = "Country",
       y = "Average Score") +
  theme_minimal() +
  theme(legend.position = "none")  # Remove legend (redundant)
```


```{r}
df %>%
  group_by(tier) %>%
  summarise(avg_score = mean(score, na.rm = TRUE)) %>%
  ggplot(aes(x = tier, y = avg_score, fill = tier)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Top Tier" = "blue", "Mid Tier" = "orange", "Low Tier" = "green")) +
  labs(title = "Average Score by University Tier",
       x = "Tier", y = "Average Score") +
  theme_minimal()

```